---
pagetitle: "Day 2"
title: | 
  | Introduction to R for Demographers
  | \vspace{1.5cm} \LARGE\emph{Workshop outline}
author: |
  | United Nations Population Division
  | 19-23 August, 2019
  | Tim Riffe
  | Max-Planck-Institute for Demographic Research
date: "20 August, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
params:
  output_dir: "../RforUNPD2019/docs"
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center}\includegraphics[trim=0 0 0 8cm, width=6cm, ]{assets/MPIDR_square_color.pdf}\\[\bigskipamount]}
- \posttitle{\end{center}}
bibliography: bibliography.bib
---

<a href="https://github.com/timriffe/RforUNPD2019" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#70B7FD; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
Today's session is about getting the most out of `R` by writing your own functions. The benefits of writing your own functions is manifold.

1. You can avoid re-deriving the same code expressions over and over. 2. Scripts using the function are shorter, which makes them easier to structure, to read, and to maintain. And that makes them more robust.
3. If you have a single copy of the function, it can get better and more robust with age: you only ever edit or fix it in one place. This avoids confusing copies of the code and needing to find them all every time you need to improve the function.
4. It encourages you to write generalizable code.
5. Functions are easier to share (and make use of when shared) than scripts.
6. There are standard ways of documenting functions.

# Reading functions
It's helpful to look at some functions to observe their structure before sitting down to write our own. Let's look at a few cases from the `DemoTools` package. For now, let's just browse functions in it on github. Here's the repository url [https://github.com/timriffe/DemoTools](https://github.com/timriffe/DemoTools)

List of some things to point out:

1. There are small helper functions
2. Functions build on each other. Some are generic pieces used in various places.
3. Optional arguments
4. Anatomy: assignment to name, arguments, body, return
5. Documentation headers
6. Beers or Sprague before / after (script to function evolution)
7. Spreadsheet to function evolution (`agesmth` examples)
8. Indentation and alignment, commenting.

This tour includes various levels of sophistication, of good and probably also bad form. Our learning today will grow from humble modularity to ambitious analysis, by combining simple utility functions in a data analysis pipeline.

# Writing functions
In this section we'll write some simple functions.
## The anatomy of a function

Here's an oldy but goody:
```{r}
hello <- function(your_name){
	paste0("Hello ", your_name, ". Have a nice day!")
}
hello("Tim")
```

1. `hello` is the name of the new function, which we know because
2. `<- function(){}` is being assigned to it, which creates functions
3. The thing in the parentheses `(your_name)` is the name of an argument to be used 
4. The part inside `{ }` is the body of the function, presumably making use of `your_name`.
5. In this particular case we're using the function `paste0()`, which concatenates text, including our argument `your_name`.

Those are the parts of a function. The function can have more arguments, which could be any kind of data of any size and shape, and the stuff that happens inside the function (inside `{ }`) should know what to do with the arguments. It should produce a result, and return it. For our little function above, it returns a character string (the last thing evaluated). In our case it printed to console because we didn't assign it to anything. We can be more explicit about that by telling it what to return:

```{r}
# Identical in this case
hello <- function(your_name){
	out <- paste0("Hello ", your_name, ". Have a nice day!")
	return(out)
}
hello("Tim")
```

So here's an outline:
```{r, eval = FALSE}
function_name <- function(argument1, argument2, ...){
	# do stuff with arguments
	return(result)
}
```

## Simplified lifetable functions
As you can see in our cheap example, functions can use other functions, as long they can be found. I think we should start straight away with an exercise. Below I give a bunch of equations for lifetables. Start with a variable `mx`, a.k.a. $m(x)$, or $\mu(x)$ if you can pretend it's continuous. We'll also need a helper-variable `ax` that tells us something about mortality within age intervals- This is one we usually assume something for, and we can be lazy and just say it's always $0.5$ for single age data.

If the mortality rate is constant in the interval, then we have some relationships that I think we can program as little functions. I think it's always a good idea to write functions with test data, so how about you use this as your test `mx` and `ax`:

```{r}
omega <- 110 # last age 
x     <- 0:omega
a     <- 0.00022
b     <- .07
mx    <- a * exp(x*b)
ax    <- rep(.5, length(mx))
```

### $q(x)$
The conditional probability of death in an age interval, $q(x)$: 
$$q(x) = \frac{m(x)}{1 + (1 - a(x)) \cdot m(x)}$$
Write a function called `mxax_to_qx()`, taking arguments `mx` and `ax` and returning a variable `qx`. Let's assume single ages for now. Note, `mx` and `ax` should be the same length (one element per age class).  Note, it's common to declare that the very last value of $q(x)$ is 1, i.e. that no one survives beyond that age.

```{r}
# mxax_to_qx <- 
```

An aside: if there's time, let's modify the function to accept an optional argument `AgeInt` that assumes intervals are equal to 1 unless otherwise told so.

### $p(x)$
The conditional survival probaility, $p(x)$
$$ p(x) = 1 - q(x) $$
Fill in the blank:
```{r}
# qx_to_px <-
```

### $l(x)$
Lifetable survivorship from birth, $l(x)$. 
$$ l(x) = \prod _{i=0}^{x-1} p(i) $$
Make the function have an optional `radix` argument, with default value equal to 1 (or something else if you prefer). Tip: there is a function called `cumprod()`.
```{r}
# px_to_lx <-
```

### $d(x)$
The lifetable death distribution, $d(x)$ is the probability at birth of dying in a given age (if the radix is equal to 1).

$$ d(x) = l(x) \cdot q(x) $$
Or you could also think of it as minus the first difference in $l(x)$
$$ d(x) = l(x) - l(x+1) $$
Can you make a function `lxqx_to_dx()`? What about `mxax_to_dx()` (make it use the previous functions!), or just `lx_to_dx()`? For this last one, note it's common if $\omega$ is nice and big to just assume that no one survives beyond that age. Remember to include a `radix` argument where appropriate. Remember also that the resulting $d(x)$ should be the same length as both inputs!
```{r}
# this one doesn't need a radix argument
# lxqx_to_dx <-

# maybe inclued AgeInt as an argument?
# mxax_to_dx <-

# this one doesn't need a radix argument
# lx_to_dx <-
```

### Lifetable exposure, $L(x)$
The total lifetable person-years lived between age $x$ and $x+1$.

$$L(x) = l(x) - (1 - a(x)) \cdot d(x)$$
write `lxax_to_dx()`
```{r}
# lxax_to_dx <-
```
Can you modify the function to include an `AgeInt` argument?

### $T(x)$
Total remaining survivorship, $T(x)$:
$$ T(x) = \sum_{i=x}^\omega L(i)$$
Try to write `Lx_to_Tx()`. This is tricky since we haven't talked about loops, but you can do it by creatively combining `rev()` (flip a vector backwards) and `cumsum()`. 

### $e(x)$
The average length of life left at each age in the lifetable, $e(x)$.
$$ e(x) = \frac{T(x)}{l(x)}$$

Can you write function `lxTx_to_ex()`? What about `mxax_to_ex()`? If you do the second one, try to make it modular, i.e. use the little functions written earlier.
```{r}
# lxTx_to_ex <-

# mxax_to_ex <-
```
What about $e(x)$ as the weighted average of $x$, where $d(x)$ are the weights? How about we write a tiny weighted mean function, `wmean()` to hold this formula?
$$ wmean = \frac{\sum x * w}{\sum w} $$
Here, $x$ is the thing we weight together and $w$ are the weights.
```{r}
# wmean <-
```
NB: insert tiny lesson on `na.rm` option.

## Summary
That was a workout! I hope that given the above repetition you'll now be able to make little utility functions. Have other favorite lifetable quantities? Have at it! Now, I suggest you take the lot of functions we just wrote and stick them in a single `R` script with an intuitive name. Put it in a new subfolder called `R` in the project folder for this workshop.

*** Time out to collect the functions we just wrote ***

# Modularity
In the previous single-purpose lifetable functions, some of the versions were already in a sense modular: A newly written function used another function that we had just written. Using this strategy to break down code into bite sized chunks enables you to structure code. I suggest that structured code enables structured thought (on methods), and vice versa-- Sometimes we start with a function that does too much and then improve it by turning functionally compact parts of it into smaller functions, a sort of whiteboard activity.

Let's look at an example of a function doing too much: The `LT` function in my `LifeTable` package, also on github:
[https://github.com/timriffe/LifeTable](https://github.com/timriffe/LifeTable)
It's not broken per se, but the fact that it isn't modular makes it difficult to maintain: There is seemingly chaotic flow control of the code using `if()` statements (to be covered tomorrow).

With this as our inspiration, let's write our own lifetable function.

## Lifetable utils in a tidy pipeline
In the first instance, let's use them in a tidy pipeline. For this, use this big HMD dataset, here:
[`HMD.rds`](https://drive.google.com/file/d/1Rc5Vb14DSglxyZ2xPaVEmEpiSCoKuC)
Stick it in a folder called `~/data`, and you can load it into your session using (this is premised on you having faithfully set up your workspace to follow or mimick the repository for this course). 
```{r}
library(here)
HMD <- readRDS(here("data","HMD.rds"))
str(HMD)
```
Take a peek at the data using `str()` or `head()` or `glimpse()` or by opening it in a data viewer. It consists in all HMD lifetable $m(x)$ and $a(x)$ values stacked, where `Country`, `Sex`, `Year`, and are key columns, and `Age`, `mx`, and `ax` are variable columns. At least for purposes of making lifetables, a practical subset consists in a chunk of data with all ages. That's why age isn't a key variable here.

As an exercise, let's fill out the lifetable columns for all unique key combinations in a tidy pipeline. To do so, we'll want to ensure all of our lifetable utility functions are loaded and available. Here's a nice trick to load them, assuming they're sitting in a script called `~/R/Functions.R` 
```{r, eval = FALSE}
source(here("R","Functions.R"))
```

Fill in the blanks (maybe there are more steps)
```{r, eval = FALSE}
HMD %>% 
  _____ %>% 
  ______
```

## Build a lifetable function
Now let's build a single lifetable function that will work in such a pipeline.

```{r}
# my_lifetable <- function(what, goes, here?){
#    what goes here?
#    return(what goes here?)
# }
```

Here arguments are given in a single `data.frame` input, but it might make sense for them to be separate vectors too. Just little things to think about when designing a function destined to be used heavily.

For your information, `DemoTools` methods have pipeline-friendly versions written in the `DDSQLtools` package, where the anticipated data chunks take the form of your `DemoData` SQL database. Pretty neat huh? `DDSQLtools` development lags a bit behind `DemoTools` development, so the collection of methods available there will grow in the coming months. In future exercises (not now), we'll use those functions.

# Documenting functions
When we toured the `DemoTools` package earlier we aw the documentation headers. These are done following some structured guidelines about what sections need to go into a function help file. These are written using a system called `roxygen2`, part of a larger movement of literate coding. The point is to keep the documentation *with* the code. This particular syntax is designed to be used in a packaging framework. I don't think we'll have time to build a package together in this visit, and it's generally a larger topic with many considerations. I suggest however that it's worth adding such headers to your self-written functions, even if they are incomplete. Doing so will help your future self, but it also makes your code more shareable. 

At a minimum, you should indicate what arguments are given in the code, using a line that follows this structure.
```{r}
#' @param arg_name description of argument here
```

Other nice sections include `#' @description` to give an overview and `#' @return` to let us know what the output is. Any other useful tips and information should be noted somehow. There's no sense in going through all the possible sections, but these are the basic ones.

***  time out to document our lifetable functions ***
Does this seem tedious to you? Well let me tell you, don't discount your own future time: reverse engineering your own code is a much bigger waste of time! If you can't figure it out the next time then you'll end up re-writing it! A major waste of time. And when you document your code other people don't need to waste their time. Let me tell you my time as been seriously wasted trying to reverse engineer thousands of lines of undocumented and unannotated code. Let's all be better people and write nicely formatted, parsimous, annotated, and documented code.

# References 
