---
pagetitle: "Day 4"
title: | 
  | Introduction to R for Demographers
  | \vspace{1.5cm} \LARGE\emph{Workshop outline}
author: |
  | United Nations Population Division
  | 19-23 August, 2019
  | Tim Riffe
  | Max-Planck-Institute for Demographic Research
date: "22 August, 2019"
output:
  html_document:
    number_sections: yes
    toc: yes
params:
  output_dir: "../RforUNPD2019/docs"
header-includes:
- \usepackage{titling}
- \pretitle{\begin{center}\includegraphics[trim=0 0 0 8cm, width=6cm, ]{assets/MPIDR_square_color.pdf}\\[\bigskipamount]}
- \posttitle{\end{center}}
bibliography: bibliography.bib
---

<a href="https://github.com/timriffe/RforUNPD2019" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#70B7FD; color:#fff; position: absolute; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# reading in data from Excel
The `readxl` package installs automatically with `tidyverse`, but you need to load it directly to use it (we'll also load a couple other helpful packages):

```{r, message = FALSE}
library(tidyverse)
library(here)
library(readxl)
```

This packge offers read functionality, and it's also platform independent, and you don't actually need Excel installed to use it. I downloaded an example dataset from Eurostat. You can download it from the repository. Open the spreadsheet so you can see you it's formatted: this one has some common formatting aspects that make it tricky to machine read, so hopefully it simulates what you might actually find.

Let's grab the table from `Sheet 1`
```{r, message = FALSE}
EURODTHS <-
        read_xlsx(here("data","eurostat.xlsx"), 
                  sheet = "Sheet 1", # sheets are straightforward,
                                     # can also index by number
                  range = "A8:X63",  # this was eyeballed
          na = ":") %>%              # each stat agency seems to prefer
                                     # different NA codes.. best to specify
  # select cols using names (character, to be sure)
  select(c("TIME", as.character(2007:2018))) %>% 
  # TIME refered to the column headers,
  # whereas GEO refers to the row names,
  # which for us are a key variable.
  rename(GEO = TIME) %>% 
  # remove the empty first row with slice()
  slice(-1) %>% 
  # finally let's get it tidy (stack the Years, which are now in columns)
  gather(key = "Year", value = "Deaths", -GEO) 
```

## Tidycells
Guess what everyone, you're lucky because a new package was released on Monday, `tidycells`. Literally this Monday, which supposedly handles the exact problem of reading strangely formatted spreadsheets and wrangling them into something sort-of tidy. Wait until it matures though, it seems. It didn't automatically handle the dataset above it seems.

## TFR and FP data wrangling
Yesterday afternoon I worked with Aisha to wrangle two spreadsheets into a single tidy gg-plottable dataset. Let's work through that from the top. To read in data, we'll use the `readxl` package also.

These are the two datasets: follow these links and click to download the data.


This exercise will introduce some new tidyverse functions

Step 1: read in both datasets, skipping columns as needed, setting column classes.

Step 2: wrange data to something tidy 

Step 3: merge the two datasets

Step 4: make a scatterplot of contraceptive use prevalence vs TFR, ideally with SDG regions colored using a pre-determined palette. 

Getting the colors will require one more data merge, since neither dataset indicated SDG. So we'll need an SDG-ISO lookup table. Then we can see how to set colors manually in `ggplot2`.

This is partially prepared and will be partially spontaneous problem-solving. So sit tight and let's see what Aisha brings in.

# Using `DemoTools`

We've looked a little bit at code from `DemoTools` to see examples for function-writing, and for use of `if` statements for conditional code execution. But we've not spoken about what's in that package. It's one of many packages that contain useful things for demographers.

## installation
This package is only on github for the time being, so to install it you'll need to use `devtools`. If there are problems for Windows users we'll have to solve them on the fly.

```{r, eval = FALSE}
# install if needed:
# install.packages("devtools")
library(devtools)
install_github("timriffe/DemoTools")
```

And now you load it as normally:
```{r}
library(DemoTools)
```
















# References 
