---
title: "Session 3"
author: "Tim Riffe"
date: "8/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `if` statements

the command `if` starts it off, then put a condition that evaluates to `TRUE` or `FALSE` in round parentheses, the contain the body of the offset code in curly braces. Either that code gets executed or not, depending on the condition in the round parentheses. Here's a hypothetical use: Do we use method A or method B? This is predicated on there having been a `method` argument that can have one value or the other.
```{r, eval = FALSE}

# la la code entering in here
if (method == "beers"){
  # then execute code that does that
}

if (method == "sprague"){
  # code that does a sprague method
} else {
  print("Sorry I don't have that method")
}
```

Use it to toggle whether to execute code or not. Say you have code that takes a long time to execute and you don't want to do it repeatedly, but rather only when necessary, and in that case
you'll manually turn it on.
```{r}
# change this to TRUE if
# we want to run this code
do.this <- FALSE
if (do.this){
HMDLT <- HMD %>% 
  group_by(Country, Year, Sex) %>% 
  LT_chunk(radix = 100000)
saveRDS(HMDLT, file = here("data","HMDLT.rds"))
}
```

Here's another example.
```{r}
x <- rnorm(10)
if (mean(x) > 0){
  print("I'm feeling lucky")
} else {
  print("Today is just not my day")
}
```

What happens if ou have more than one `TRUE` / `FALSE`? Then it takes the first one and warns you:
```{r}
if (x > 0){
  print("I'm feeling lucky")
}
```

Here are some ways to do logical checking on vectors.
```{r}
any(x > 0)
all(x > 0)
all(! (x > 0))
```

## what about lots of little conditions?
The case of recoding `ifelse()` is vectorized, meaning it'll do a lot at once without you telling it so explicitly. This works well when the number of cases to deal with is smallish.

```{r}
edu <- c("sec","pri","ter","ter","sec","pri","pri")
ifelse(edu == "pri", "pri" , ">pri" )
# double-nested
ifelse(edu == "pri", 1,
       ifelse(edu == "sec", 2, 3))

# mini example of vectorization
a <- rnorm(10)
b <- rnorm(10)
a+b
```

But what if there are lots of cases? Do we really need to nest that deep? That's ugly. Here's a way around it with `case_when()` from the tidyverse.

```{r, message = FALSE}
library(tidyverse)
x <- 1:50
x -  x %% 5

case_when(
  (x %% 5 == 0) & (x %% 7 == 0) ~ "fizzbuzz",
  x %% 5 == 0 ~ "fizz",
  x %% 7 == 0 ~ "buzz",
  TRUE ~ as.character(x)
)
```

Final case, for when the number of things you need to recode gets really big. Imagine something like reducing 1000 occupational codes to 10, or adding a Continent variable to dataset if you have country codes plus a lookup table. This trick uses named indexing to do its work. It makes sense if there are no `NA`s.

This will also show named indexing.
```{r}
x <- 1:26
names(x) <- letters

# the values of this vector should be the destination codes.
recode_vector <- sample(1:4, 26, replace = TRUE)

# the names of the vector are the origin codes. Be thorough!
names(recode_vector) <- letters

# you can use the recode_vector like this:
FakeData <- data.frame(ID = sample(letters, 
                       100, 
                       replace = TRUE))

# Use the column containing the origin codes to
# 'select' from the recode_vector. It expands, 
# repeats, and rearranges as required, neat!
FakeData$ID2 <- recode_vector[FakeData$ID]
head(FakeData)
```

Another alternative still would be to just get the two data objects, the main one and the lookup table, and merge them as a data operation. For tips on this (and more `dplyr` functions), look at the `dplyr` cheat sheet.

Logical operators that I can think of right now.
```{r}
"a" %in% letters
c("a","Z") %in% letters

"a" == letters
c("a","Z") == letters

x <- rnorm(10)
x >= 0
# ==
# !=
  
```

# iteration
Well, not iterartion in general, just for-loops for now. Looping is the most general solution to repetetive code execution. It's done almost the same way in many languages.

```{r, eval = FALSE}
for (yr in years){
  # repeat this code
}
```

```{r}
countries <- c("AUT", "BEL", "DEU", "NLD", "GBR", "USA")
for (cntry in countries){
  print(cntry)
}

```

1. Rule 1, you may or may not actually use the iterator. Here's a funny example. We increment a counter (for no reason at the moment) to illustrate that the loop repeats once for each element in set that is iterated over.

```{r}
i <- 0
for (cntry in countries){
  i <- i + 1
  print(i)
}
```

2. The body knows about objects that were created before the loop. It's kind of a trivial property, and we point this out because the body of the loop is offset but it is still in your environment at the same level.

```{r}
A <- 5
B <- 6
for (i in 1:10){
  print(A / i + B * i)
}
```

You can use loops for trivial things.

```{r}
x <- rnorm(100)
y <- rnorm(100)
xi <- x * 0
yi <- y * 0
for (i in 1:length(x)){
  xi[i] <- sum(x[1:i])
  yi[i] <- sum(y[1:i])
}
```

In the above we pre-allocated the container for the results, which is best practice. But we don't need to. We could also grow an object as the loop progresses:
```{r}

xi <- 0
yi <- 0
for (i in 1:length(x)){
  xi[i] <- sum(x[1:i])
  yi[i] <- sum(y[1:i])
}
# R grows your vector to the required
# size without complaining, dangerously polite
xi[1000] <- 1
```

Many times there is a simple function that already does what we need.
```{r, eval = FALSE}
cumsum(x) == xi
```

## Example using the HMD data:

```{r}
library(here)
library(tidyverse)
source(here("R","Functions.R"))
HMD <- readRDS(here("data","HMD.rds"))

HMDLT <- HMD %>% 
  group_by(Country, Year, Sex) %>% 
  LT_chunk(radix = 100000)
```

What would this look like in a loop?

```{r, eval = FALSE}
countries <- unique(HMD$Country)
sexes     <- unique(HMD$Sex)

HMDLT     <- list()
i         <- 0
for (ctry in countries){
  for (sex in sexes){
    # calculate unique years for
    # this subset without needing
    # to allocate the entire subset
    years <- 
      filter(HMD,
             Country == ctry,
             Sex == sex) %>% 
      # this is the new function.
      # it extracts a column, kind of
      # like $
      pull(Year) %>% 
      unique()
    # This is what we can avoid:
    # chunktoobig <- filter(HMD,
    #          Country == ctry,
    #          Sex == sex)
    # years <- unique(chunktoobig$Year)
    for (yr in years){
      i <- i + 1
      chunk <- filter(HMD,
                      Country == ctry,
                      Sex == sex,
                      Year == yr)
      HMDLT[[i]] <- LT_chunk(chunk, 
                             radix = 100000)
    } # end year loop
  } # end sex loop
} # end country loop
HMDLT <- bind_rows(HMDLT)
```

Lesson: don't grow objects in loops. It's the memory reallocation stuff under the hood that will slow you down. But there are alternatives.

```{r}

# make a list, where each element is
# a subset
HMDLT <- split(HMD, list(HMD$Country, 
                         HMD$Sex,
                         HMD$Year
                ),
               drop = TRUE)
HMDLT <- lapply(HMDLT, 
                LT_chunk, 
                radix = 100000)

HMDLT <- bind_rows(HMDLT)
```

But we can borrow a tip from this strategy and not abandon loops altogether:

```{r, eval = FALSE}
HMDLT <- split(HMD, list(HMD$Country, 
                         HMD$Sex,
                         HMD$Year
                ),
               drop = TRUE)
# for-loop goes here:
# here we iterate down the elements
# of this largish list
for (i in 1:length(HMDLT)){
  HMDLT[[i]] <- LT_chunk(HMDLT[[i]], 
                         radix = 100000)
}
HMDLT <- bind_rows(HMDLT)
```

Defensive coding inside this last loop technique:

```{r, eval = FALSE}
HMDLT <- split(HMD, list(HMD$Country, 
                         HMD$Sex,
                         HMD$Year
                ),
               drop = TRUE)
# for-loop goes here:
# here we iterate down the elements
# of this largish list
for (i in 1:length(HMDLT)){
  if (i == 100){
    HMDLT[[i]]$ax <- as.character(HMDLT[[i]]$ax)
  }
  HMDLT[[i]] <- 
    try(LT_chunk(HMDLT[[i]], 
                         radix = 100000))
}
HMDLT[[100]] # here's the error message. 
# One could also, instead of keeping the message,
# capture some more info and assign that instead.
HMDLT[[101]] %>% head()
# would need to remove the error list elements first.
HMDLT <- bind_rows(HMDLT)
```

# Exercises

1. Start with an integer.
2. i) if it is even, divide by 2
  ii) if it is odd, multiply by 3 and add 1
3. wash-wrinse-repeat, count how many steps it takes. That's the Collatz number.

Step 1, write some lines of code that finds the number. Some correct answers: 1 has a Collatz number of 0. 2 is 1. 3 is 7 (did it on board).
```{r}
number <- 3
Cnumber <- 0

for (i in 1:1000){
  if (number == 1){
    break
  }
  Cnumber <- Cnumber + 1
  if (number %% 2 == 0){
    number <- number / 2
  } else {
    number <- number * 3 + 1
  }
  number <- as.integer(number)
}
```

Step 2, turn it into a function

```{r}
Collatz <- function(number, maxit = 1e4){
  Cnumber <- 0
  for (i in 1:maxit){
    number <- as.integer(number)
    if (number == 1){
      return(Cnumber)
    }
    Cnumber <- Cnumber + 1
    if (number %% 2 == 0){
      number <- number / 2
    } else {
      number <- number * 3 + 1
    }
  }
  warning("you better increase maxit!\nNo Cmnumber for you!")
}
```


Step 3, let's look at the neat pattern-not-pattern

```{r}

numbers <- 1:10000
Cnumbers <- numbers * 0
for (n in numbers){
  Cnumbers[n] <- Collatz(n)
}
data.frame(n = numbers, 
           Coll = Cnumbers) %>% 
  ggplot(mapping = aes(x = n, 
                       y = Coll)) +
  geom_point() +
  scale_x_log10()

```

## Coale's iterative method for finding intrinsic $r$

First let's get a an $L(x)$ vector and an $F(x)$ (ASFR) vector to play with.
```{r}
HFD <- readRDS(here("data","HFD.rds"))
HMD <- readRDS(here("data","HMD.rds"))
Lx <- HMD %>% 
  group_by(Country, Sex, Year) %>% 
  LT_chunk(radix = 100000) %>% 
  filter(Country == "SWE",Year == 1800,
         Sex == "f") %>% 
  pull(Lx)
Lx <- Lx / 100000

Fx <- HFD %>% 
  filter(Country == "SWE",Year == 1900) %>% 
  pull(ASFR)
Fx <- c(rep(0,11), Fx, rep(0,56)) * .4886 # discount for female births
```


Steps:
1. Calculate $R(0)$ (aka NRR) as:
 $$ R(0) = \sum L(x) F(x)$$
2. Assume $T$ (mean length of generation in the stable population) is something like 29.
3. calcualte a first guess at $r$
$$ r^i = \frac{log(R(0))}{T}$$
4. now in a loop, update `r` in using two steps
   (i) get a residual
   $$ \delta = \left(\sum e^{-r^i * x}F(x)*.4886*L(x)\right) - 1 $$
   (ii) now use $\delta$ to update $r^i$:
   $$ r^{i+1} = r^i + \frac{\delta}{T - \frac{\delta}{r^i}}$$
5. When $\delta = 0$ you're done!

And here's that code!
```{r}
x     <- .5:110.5
maxit <- 100
tol   <- 1e-10

Coale_r <- function(Fxf, Lx, 
                    maxit = 100, tol = 1e-12){
  R0    <- sum(Lx * Fxf) # fixed quantity
  G     <- 29            # assumed
  ri    <- log(R0) / G   # first guess
  for (i in 1:maxit){
    # get residual
    delta <- sum(exp(-ri * x) * Fxf * Lx) - 1
    # use it to adjust r estimate
    ri    <- ri + delta / (G - delta / ri)
    # check for convergence
    if (delta^2 < tol){
      break
    }
  }
  return(ri)
}
# messing around to look for how fert would need to change
# to achieve stationarity.
Coale_r(Fx*1.0252, Lx)

```


























